The role of cross-linguistic lexical similarity on bilingual word acquisition

### 2022-09-23 manuscript/manuscript.Rmd
Fix random slopes and update manuscript

---
title             : "The role of cross-linguistic lexical similarity on bilingual word acquisition"
shorttitle        : "Cross-linguistic similarity and  word acquisition"

author: 
  - name          : "Gonzalo Garcia-Castro"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Ramon Trias Fargas, 25-27, 08005 Barcelona, Spain"
    email         : "gonzalo.garciadecastro@upf.edu"
  - name          : "Daniela Avila-Varela"
    affiliation   : "1"
  - name          : "Ignacio Castillejo"
    affiliation   : "2"
  - name          : "Nuria Sebastian-Galles"
    affiliation   : "1"

affiliation:
  - id            : "1"
    institution   : "Center for Brain and Cognition, Universitat Pompeu Fabra, Barcelona, Spain"
  - id            : "2"
    institution   : "Universidad Autónoma de Madrid"

authornote: |
  Campus de Ciutadella, Universitat Pompeu Fabra, 08005, Barcelona, Spain

abstract: |
  Bilinguals face the challenging task of learning words from languages with overlapping phonologies. Floccia et al. (2018) reported larger vocabulary sizes for 24-month-old bilinguals that were learning languages that shared a greater amount of cognates (e.g., English-Dutch). The mechanisms underlying this effect remain unknown. We explore two compatible scenarios. First, we test whether cognates are learnt earlier than non-cognates. This would account for the difference in vocabulary size associated to the amount of shared cognates across languages. Second, we explore the possibility that the word-forms of one language interact with those form the other language, scaffolding the acquisition of their translation equivalents when their phonologies overlap. This mechanism, in line with the parallel activation account of bilingual speech perception, would provide a plausible explanation to why cognates are acquired ealier by bilinguals. We developed an online tool to collect parental reports of receptive and productive vocabularies from children learning Catalan and/or Spanish, and present data on receptive and productive vocabulary of bilingual toddlers aged 12 to 34 months.

  <!-- https://tinyurl.com/ybremelq -->

keywords          : "lexical acquisition, vocabulary, bilingualism"
wordcount         : "X"

bibliography      : ["references.bib"]

floatsintext      : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
keep_text         : false
class             : "man"

documentclass     : "apa6"
classoption       : "man"
output            :
  word_document:
    reference_docx: "manuscript_template.docx"

---

@@

# load packages
library(bayesplot)
library(brms) 
library(conflicted)
library(dplyr) 
library(forcats)
library(ggplot2) 
library(ggsci)
library(gt) 
library(here) 
library(janitor)
library(knitr)
library(lubridate)
library(multilex)
library(patchwork) 
library(papaja) 
library(purrr)
library(quarto)
library(rmarkdown)
library(scales) 
library(stringdist)
library(stringr)
library(tidybayes) 
library(tidyr)

# load and process objects
tar_load_everything()
participants <- participants %>% 
    left_join(select(multilex_data$logs, id, time, edu_parent))
items <- items %>% 
    left_join(select(pool, item, language, category, class))

@@

Before the end of their first year of life, infants start directing their gaze to some objects when hearing their labels, according to both experimental data [@jusczyk1995infants; @tincoff1999some; @bergelson2012months; @bergelson2015early] and parental reports [e.g., @fenson2007macarthur]. During the last half of their second year, they acquire new words at an increasingly fast rate [@goldfield1990early; @fenson1994variability; @mayor2011statistical; @bloom2002children; @bergelson2020comprehension]. These early stages of lexical acquisition are characterised by substantial variation between children, reflected, for instance, on the variation on the number of words they know [i.e. vocabulary size; @fenson1994variability; @frank2021variability] or on the proportion of those words that fall into the category of nouns, as opposed to verbs, adjectives, or function words [e.g., @nelson1973structure, bates1994developmental].

Despite this variability, children's trajectories of vocabulary growth seem quite stable across languages. @tardif2008baby collected data about the first ten words acquired by 10 to 16 month-old infants living in the United States, Hong Kong, and Beijing. Since birth, these infants had been learning English, Cantonese and Mandarin, respectively. The authors found a common pattern across the three groups: their first ten words referred to roughly the same concepts, namely relatives/caretakers (*daddy*, *mommy*), social routines (*bye*, *uh-oh*) or animals (*woof-woof*). These results were later extended by @frank2021variability to a wider diversity languages, who also reported that such cross-language commonalities are stronger at earlier stages of lexical acquisition, compared to later stages. Most of the literature on early word acquisition, however, has been conducted on monolingual children, and neglects the problem of how bilinguals--who represent a substantial proportion of the population in most societies--acquire words at early ages.

There is evidence that bilinguals know less words in each of their languages than monolinguals, but also that both groups know a similar amount of words when the two languages are aggregated. For example, @hoff2012dual found that English-Spanish bilingual toddlers in South Florida knew less words in English than monolinguals, who only learnt English. Both groups knew a similar total amount of words when both English and Spanish vocabularies were counted together, pointing to the importance of collecting data on both languages when assessing bilinguals' communicative development. Other studies have provided converging evidence that bilinguals know a similar--or even larger--number of words than monolinguals, only when the languages are aggregated [@oller2002language; @pearson1994patterns; @pearson1993lexical; @patterson2004comparing; @patterson2004bilingual; @smithson2014bilingualism; @petitto2001bilingual; @gonzalez2020bilingual]. While these studies have mostly relied on samples of bilingual children learning two relatively distant languages, as it is the case of English and Spanish, it is unclear whether children learning typologically more similar languages also know less words in each of their languages than monolinguals. What role could linguistic distance play during early vocabulary growth?

@@

Floccia et al. pointed to *parallel activation* as the main mechanism underpinning their results. The parallel activation hypothesis suggests that bilinguals activate both languages simultaneously during speech production or comprehension. This phenomenon is the result of the activation of lexical representations in both languages, even when only one is in use during production [@costa2000cognate; @hoshino2008cognate] or comprehension [@thierry2007brain]. One of the clearest examples of parallel activation was provided by @costa2000cognate. In this study, Catalan-Spanish monolingual and bilingual adults were asked to name pictures of common objects in Spanish. In half of the trials, the object labels were cognates in Spanish and Catalan (*árbol*-*arbre*, translations of *tree*), whereas in the other half of the trials labels were non-cognates (*mesa*-*taula*, translations of *table*). Bilinguals named cognate pictures faster than non-cognate pictures, even after adjusting for the lexical frequency of the items. Importantly, Spanish monolinguals did not show this effect. These results suggest that, for the bilinguals, Catalan phonology was activated during the production of Spanish words, facilitating the naming of cognate pictures. Several recent studies have also provided similar evidence in comprehension in children [e.g., @von2012language; @poulin2013lexical]. Parallel activation is therefore a plausible mechanism to account for Floccia et al.'s results: cognates increase the amount of activation in both languages, facilitating word acquisition, and ultimately leading to children learning language pairs with a larger proportion of cognates are predicted to show larger vocabulary sizes. A missing piece in this account is how increase parallel activation might facilitate the acquisition new words.

In this study, we hypothesise that cognates are acquired earlier in age than non-cognates, which might explain why children learning languages that share many cognates show larger vocabulary sizes. Evidence supporting this claim is scarce. To the best of our knowledge, only two studies have provided direct data on this issue. First, @bosch2014first used vocabulary parental reports (152 lexical items) from 48 Catalan-Spanish bilinguals aged 18 months, and found that cognates represented a larger proportion of participant's vocabulary than non-cognates. Second, @schelletter2002effect reported a longitudinal single case of one English-German bilingual who produced cognates earlier than non-cognates, on average. The low sample size in these studies makes it challenging to draw strong conclusions about the effect of cognateness on vocabulary growth. Floccia et al.'s estimates are statically more reliable given their (much larger) sample size, but their study was not aimed at testing the effect of cognateness on age of acquisition directly. In their discussion the authors state the following (pp. 70):

@@

We identify two main reasons why an earlier age of acquisition for cognates than for non-cognates is an unwarranted conclusion from Floccia et al.'s results. First, the response variable used was the proportion of words each participant understood and/or produced (i.e., vocabulary size), from the list of lexical items in the vocabulary checklists. By aggregating the responses from all items into a single datum per child, information about the acquisition status of cognates vs. non-cognates was lost. Second, all participants were aged ~24 months, meaning that even if the unaggregated responses to individual items were included as response variable, the possible effect of cognateness could only be interpreted as an increase or decrease in the likelihood of participants at such age to have acquired each item, and not as an increase or decrease in the age of acquisition of such item. In summary, the current evidence supporting an earlier age of acquisition of cognates vs. non-cognates presents some limitations that prevents drawing sound conclusions about this effect.

Our second hypothesis detaches from the fact that, for parallel activation to take place, and for cognateness to facilitate the acquisition of a word form in a given language, the child must be familiar with the (phonologically similar) translation of the word, in the other language. If the child is not previously familiar with one of the word forms, the acquisition of its translation in  language cannot be facilitated by cognateness. This implies that, if cognateness facilitates the acquisition of a word form, it can only do so after the child has acquired one of the word forms of the translation pair. Given that children are more likely to acquire words from languages to which they are exposed more often [@david2008individual; @cattani2014much; @thordardottir2006bilingual], the acquisition of words in the language of lower exposure should, on average, be more susceptible to the effect of cognateness. Therefore, we hypothesised that the effect of cognateness for words in the language of lower exposure would be larger than in the language of higher exposure.

In line with this hypothesis, the vocabulary size associated with linguistic similarity that Floccia et al. reported was larger in the additional language vocabulary than in English vocabulary. Most participants in their sample were English-dominant, meaning that their relative amount of exposure to English was larger than in the additional language. Therefore, participants may have, on average, learnt the English word-form of translation equivalents earlier than the word-form in the additional language.  If this is the case, then the acquisition of English words by English-dominant participants would rarely benefit from their cognate status (the other word-form is not available yet), while the acquisition of words in the additional language would benefit from their phonological similarity with the (available) English form. To test this hypothesis, we conditioned the effect of cognateness to the relative degree of exposure to the language.

In summary, we investigated the role of cognateness in bilingual word acquisition (conditional to the degree of relative exposure to the language) as possible mechanism to account for the facilitation  of language similarity on bilingual children's vocabulary size. We hypothesised that cognate words would be acquired earlier than non-cognate words, and that this effect should be larger for words in the language with lower exposure. Using an online vocabulary checklist--carefully designed specifically for this study--we collected data from a sample of children aged `r min(floor(participants$age))` to `r max(floor(participants$age))` months learning Catalan and/or Spanish, with varying degrees of exposure to each language. We then modelled the probability of participants being reported by their parents to understand and/or say each word in the checklist, conditional to its cognate status in Catalan and Spanish and participants' degree of exposure to the corresponding language, while adjusting for participants' age and the item's lexical frequency.

@@

We collected data from `r printnum(n_participants)` children from the Metropolitan Area of Barcelona between `r format(dates[1], "%dth %B, %Y")` and `r format(dates[2], "%dth %B, %Y")`. All families gave informed consent before participating and this study was approved by the Comitè d'Ètica de la Investigació amb Medicaments (CEIm) from Hospital del Mar (Barcelona, Spain), code XXXXXXXXX. Families were contacted by e-mail or phone if their child were aged between 10 and 34 months, and had not previously reported to be exposed more than 10% of the time to a language other than Spanish or Catalan. Upon consent, families were sent a link to the questionnaire via e-mail, which they filled from a computer, laptop, or mobile device in a browser within the two week following the invitation to participate. Table 1 summarises the distribution of participants across ages and degrees of exposure (DoE) to the non-dominant language (i.e., their degree of bilingualism).

```{r participants-lp, message=FALSE, warning=FALSE, paged.print=FALSE, fig.width=7, fig.cap="Participant sample size by age and degree of exposure to a second language. Information about DoE was provided by families when filling the questionnaire. For illustration purposes this table, DoEs were binned into 10\\% wide bins, and ages (in months) were round to the nearest integer. Hyphens indicate that no participants from that specific combination of age filled the questionnaire."}
participants %>% 
    mutate_at(
        vars(starts_with("doe_")), 
        ~floor(.*10)/10
    ) %>% 
    mutate(age = cut_width(floor(age), 4)) %>% 
    count(age, doe_catalan) %>% 
    arrange(-doe_catalan) %>% 
    pivot_wider(names_from = age, values_from = n) %>% 
    gt() %>% 
    fmt_missing(everything(), everything(), missing_text = "--") %>% 
    cols_label(
        doe_catalan = "DoE Catalan"
    ) %>%
    fmt_percent("doe_catalan", decimals = 0) %>% 
    tab_spanner("Age (months)", 2:8)
```

@@

The questionnaire was implemented on-line using the formR platform [@arslan2020formr], and was structured in three blocks: a (1) language questionnaire, a (2) demographic survey, and a (3) Catalan and a Spanish vocabulary checklists. Vocabulary checklists followed a similar structure as the Oxford Communicative Developmental Inventory [@hamilton2000infant] and consisted in two lists of words: one in Catalan and one in Spanish. The Catalan inventory contained `r n_item_language["catalan"]` items and the Spanish inventory contained `r n_item_language["spanish"]`. Items in one language were translation equivalents of the items in the other language (e.g., whenever *gos* [dog] was included in the Catalan inventory, the word *perro* was included in the Spanish inventory), roughly following a one-to-one mapping. In case two translation equivalents were possible for a given word, both were inluded as separate items (e.g., Catalan *acabar* [*to finish*] and Spanish *acabar* and *terminar*), or merged them into a single item (e.g., Spanish *mono* [*monkey*] and Catalan *mono/mico*). We included items from a diverse sample of `r n_categories` semantic/functional categories (see Appendix 1). For the analyses included in this study, we excluded items from the adverbs, auxiliary words, connectives, interjections and games and routines categories, so that only data from content words (nouns, adjectives, and verbs) were used.

@@

```{r item-summary, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE, fig.width=7, fig.cap="Lexical frequencies. Mean, standard error, and 95\\% confidence interval of lexical frequencies of items included in the Catalan and Spanish lists, reported separately for identical cognates, non-identical cognates, and non-cognates."}
items %>%
    unnest(list) %>% 
    group_by(language, list) %>% 
    summarise_at(
        vars(freq, n_phon, lv), 
        lst(n = ~sum(!is.na(.)), mean, sd, min, max)
    ) %>% 
    select(-c(n_phon_n, lv_n)) %>% 
    rename(n = freq_n) %>% 
    gt(groupname_col = "language") %>% 
    fmt_number(matches("freq_|n_phon_mean|n_phon_sd")) %>% 
    fmt_percent(matches("lv_")) %>% 
    tab_spanner("Frequency (Zipf)", columns = starts_with("freq_")) %>% 
    tab_spanner("# Phonemes", columns = starts_with("n_phon_")) %>% 
    tab_spanner("Levenshtein", columns = starts_with("lv_")) %>% 
    cols_merge_range(col_begin = "freq_min", col_end = "freq_max") %>% 
    cols_merge_range(col_begin = "n_phon_min", col_end = "n_phon_max") %>% 
    cols_merge_range(col_begin = "lv_min", col_end = "lv_max") %>% 
    cols_label(
        list = "",
        n = md("*N*"),
        freq_mean = md("Mean"),
        freq_sd = md("*SD*"),
        freq_min = md("Range"),
        n_phon_mean = md("Mean"),
        n_phon_sd = md("*SD*"),
        n_phon_min = md("Range"),
        lv_mean = md("Mean"),
        lv_sd = md("*SD*"),
        lv_min = md("Range")
    ) %>% 
    tab_style(
        cell_text(style = "italic"),
        cells_column_labels(everything())
    ) %>% 
    tab_style(
        cell_text(weight = "bold"),
        cells_column_spanners(everything())
    )
```

@@

```{r coefs, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
coefs <- as.data.frame(fixef(model_fit_7)) %>% 
    clean_names()

draws_fixed <- gather_draws(model_fit_7, `b_.*`, regex = TRUE)
```

Table 2 shows the summary of the posterior distribution of the fixed coefficients of the model.

```{r results-fixed, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
str_repl <- c(
    "b_age_std:doe_std:lv_std" = "Age \u00d7 DoE \u00d7 Levenshtein",
    "b_age_std:doe_std" = "Age \u00d7 DoE",
    "b_age_std:lv_std" = "Age \u00d7 Levenshtein",
    "b_doe_std:lv_std" = "DoE \u00d7 Levenshtein",
    "b_lv_std" = "Levenshtein (+1 SD)",
    "b_doe_std" = "DoE (+10%)",
    "b_n_phon_std" = "# Phonemes (+1 SD)",
    "b_freq_std" = "Frequency (+1 SD)",
    "b_age_std" = "Age (+1 month)",
    "b_Intercept[1]" = "Intercept (Production)",
    "b_Intercept[2]" = "Intercept (Comprehension)"
)

>>>>

---
title: "The role of cross-linguistic lexical similarity on bilingual word acquisition"
author:
- name: Gonzalo Garcia-Castro
  affiliation: '1'
  corresponding: yes
  address: Ramon Trias Fargas, 25-27, 08005 Barcelona, Spain
  email: gonzalo.garciadecastro@upf.edu
- name: Daniela Avila-Varela
  affiliation: '1'
- name: Ignacio Castillejo
  affiliation: '2'
- name: Nuria Sebastian-Galles
  affiliation: '1'
shorttitle: Cross-linguistic similarity and  word acquisition
output:
  pdf_document: default
  word_document:
    reference_docx: manuscript_template.docx
authornote: |
  Campus de Ciutadella, Universitat Pompeu Fabra, 08005, Barcelona, Spain
abstract: |
  Bilinguals face the challenging task of learning words from languages with overlapping phonologies. Floccia et al. (2018) reported larger vocabulary sizes for 24-month-old bilinguals that were learning languages that shared a greater amount of cognates (e.g., English-Dutch). The mechanisms underlying this effect remain unknown. We explore two compatible scenarios. First, we test whether cognates are learnt earlier than non-cognates. This would account for the difference in vocabulary size associated to the amount of shared cognates across languages. Second, we explore the possibility that the word-forms of one language interact with those form the other language, scaffolding the acquisition of their translation equivalents when their phonologies overlap. This mechanism, in line with the parallel activation account of bilingual speech perception, would provide a plausible explanation to why cognates are acquired ealier by bilinguals. We developed an online tool to collect parental reports of receptive and productive vocabularies from children learning Catalan and/or Spanish, and present data on receptive and productive vocabulary of bilingual toddlers aged 12 to 34 months.
  <!-- https://tinyurl.com/ybremelq -->
keywords: lexical acquisition, vocabulary, bilingualism
wordcount: X
bibliography: references.bib
floatsintext: yes
figurelist: no
tablelist: no
footnotelist: no
linenumbers: yes
mask: no
draft: no
keep_text: no
class: man
documentclass: apa6
classoption: man
affiliation:
- id: '1'
  institution: Center for Brain and Cognition, Universitat Pompeu Fabra, Barcelona,
    Spain
- id: '2'
  institution: Universidad Autónoma de Madrid
---

@@

# load packages
library(brms) 
library(conflicted)
library(dplyr) 
library(forcats)
library(ggplot2) 
library(ggsci)
library(gt) 
library(here) 
library(janitor)
library(knitr)
library(lubridate)
library(multilex)
library(patchwork) 
library(papaja) 
library(purrr)
library(scales) 
library(stringr)
library(tidybayes) 
library(tidyr)

# load and process objects
tar_load_everything()

participants <- participants %>% 
    left_join(select(multilex_data$logs, id, time, edu_parent))
items <- items %>% 
    left_join(select(pool, item, language, category, class))

@@

The foundations of word learning are in place early in age: before the end of their first year of life, infants start directing their gaze to some objects when hearing their labels, according to both experimental data [@jusczyk1995infants; @tincoff1999some; @bergelson2012months; @bergelson2015early] and parental reports [e.g., @fenson2007macarthur]. During the last half of their second year, they acquire new words at an increasingly fast rate [@goldfield1990early; @fenson1994variability; @mayor2011statistical; @bloom2002children; @bergelson2020comprehension]. These early stages of lexical acquisition are characterised by substantial variation across children, reflected, for instance, on the variation on the number of words they know [i.e. vocabulary size; @fenson1994variability; @frank2021variability] or on the proportion of those words that fall into the category of nouns, as opposed to verbs, adjectives, or function words [e.g., @nelson1973structure, bates1994developmental].

Despite this variability, children's trajectories of vocabulary growth seem quite stable across languages. @tardif2008baby collected data about the first ten words acquired by 10 to 16 month-old infants living in the United States, Hong Kong, and Beijing. Since birth, these infants had been learning English, Cantonese and Mandarin, respectively. The authors found a common pattern across the three groups: their first ten words referred to roughly the same concepts, namely relatives/caretakers (*daddy*, *mommy*), social routines (*bye*, *uh-oh*) or animals (*woof-woof*). These results were later extended by @frank2021variability to a wider diversity languages, also reporting that such cross-language commonalities are stronger at earlier stages of lexical acquisition, compared to later stages. Most of the literature on early word acquisition, however, has been conducted on monolingual children, and neglects the problem of how bilinguals--who represent a substantial proportion of the population in most societies--acquire words at early ages.

There is evidence that bilinguals know, on average, less words in each of their languages than monolinguals, but also that both groups know a similar amount of words when the two languages are aggregated. For example, @hoff2012dual found that English-Spanish bilingual toddlers in South Florida knew less words in English than monolinguals, who only learnt English. Both groups knew a similar total amount of words when both English and Spanish vocabularies were counted together, pointing to the importance of collecting data on both languages when assessing bilinguals' communicative development. Other studies have provided converging evidence that bilinguals know a similar--or even larger--number of words than monolinguals, only when the languages are aggregated [@oller2002language; @pearson1994patterns; @pearson1993lexical; @patterson2004comparing; @patterson2004bilingual; @smithson2014bilingualism; @petitto2001bilingual; @gonzalez2020bilingual]. While these studies have mostly relied on samples of bilingual children learning two relatively distant languages, as it is the case of English and Spanish, it is unclear whether children learning typologically more similar languages also know less words in each of their languages than monolinguals. What role could linguistic distance play during early vocabulary growth?

@@

Floccia et al. pointed to *parallel activation* as the main mechanism underpinning their results. The parallel activation hypothesis suggests that bilinguals activate both languages simultaneously during speech production or comprehension, and that this phenomenon is the result of the activation of lexical representations in both languages, even when only one is in use during production [@costa2000cognate; @hoshino2008cognate] or comprehension [@spivey1999cross; @thierry2007brain]. One of the clearest examples of parallel activation was provided by @costa2000cognate. In this study, Catalan-Spanish monolingual and bilingual adults were asked to name pictures of common objects in Spanish. In half of the trials, the object labels were cognates in Spanish and Catalan (*árbol*-*arbre*, translations of *tree*), whereas in the other half of the trials labels were non-cognates (*mesa*-*taula*, translations of *table*). Bilinguals named cognate pictures faster than non-cognate pictures, even after adjusting for the lexical frequency of the items. Importantly, Spanish monolinguals--unfamiliar with the Catalan translations of the Spanish words they uttered--did not show this effect. These results suggest that bilinguals' Catalan phonology was activated during the production of Spanish words, facilitating the naming of cognate pictures. Several subsequent studies have also provided similar evidence in comprehension in children [e.g., @von2012language; @poulin2013lexical]. Parallel activation is therefore a plausible mechanism to account for Floccia et al.'s results: cognates increase the amount of activation in both languages, facilitating word acquisition, and ultimately leading to children learning language pairs with a larger proportion of cognates are predicted to show larger vocabulary sizes. A missing piece in this account is how increase parallel activation might facilitate the acquisition new words.

In this study, we hypothesise that cognates are acquired earlier in age than non-cognates, which might explain why children learning languages that share many cognates show larger vocabulary sizes. Evidence supporting this claim is scarce. @bosch2014first used vocabulary parental reports (152 lexical items) from 48 Catalan-Spanish bilinguals aged 18 months, and found that cognates represented a larger proportion of participant's vocabulary than non-cognates. @schelletter2002effect reported a longitudinal single case of one English-German bilingual who produced cognates earlier than non-cognates, on average. The low sample size in these two studies makes it challenging to draw strong conclusions about the effect of cognateness on vocabulary growth. On the other hand, Floccia et al.'s estimates are statically more reliable given their (much larger) sample size, but their study was not aimed at testing the effect of cognateness on age of acquisition directly. In their discussion the authors state the following (pp. 70):

@@

We identify two main reasons why an earlier age of acquisition for cognates than for non-cognates is an unwarranted conclusion from Floccia et al.'s results. First, the response variable used was the proportion of words each participant understood and/or produced (i.e., vocabulary size), from the list of lexical items in the vocabulary checklists. By aggregating the responses from all items into a single datum per child, information about the acquisition status of cognates vs. non-cognates was lost. Second, all participants were aged ~24 months, meaning that even if the unaggregated responses to individual items were included as response variable, the possible effect of cognateness could only be interpreted as an increase or decrease in the likelihood of participants at such age to have acquired each item, and not as an increase or decrease in the age of acquisition of such item.

More recently, @mitchell2022cognates addressed this issue overcoming some of these pitfalls. Using a larger, longitudinal sample of 47 16-to-30 month-old French-English bilinguals living in Canada, the authors collected data on expressive vocabulary data in both languages. They created two lists of translation equivalents: one made of 131 cognates, and one made of 406 non-cognates. The proportion of translation equivalents that children were reportedly able to produce was higher in the cognate lists than in the non-cognate list across ages, even when both lists were matched by semantic category (furniture, animals, food were similarity represented in both lists) and age of acquisition norms (an index of word difficulty). These findings shed some light on the ongoing exploration of why (if at all) bilinguals' vocabulary size grows faster when both languages are phonologically more similar: word-forms sharing more phonemes with their translation equivalents (i.e., cognates) seem to be acquired faster.

As @mitchell2022cognates note in their discussion, two competing hypotheses remain untested. The facilitative effect of cognateness can be explained by, at least, two alternative, arguably incompatible, mechanisms. On the one hand, cognateness might operate as a facilitator of the acquisition of both forms of a translation pair, with every exposure to either form activating the other, and therefore strengthening the association between both and their referent. If this is the case, then cognateness should increase the probability of acquisition of the forms in both languages equally.

Another possibility is that cognateness plays a role in the acquisition of a translation pair only when at least one of the word-forms has been acquired. This account departs from the premise that, if the child is not previously familiar with one of the word forms, the acquisition of its translation in the other language cannot be facilitated by cognateness. This implies that, if cognateness facilitates the acquisition of a word form, it can only do so after the child has acquired one of the word forms of the translation pair. Given that children are more likely to acquire words from languages to which they are exposed more often [@david2008individual; @cattani2014much; @thordardottir2006bilingual], the acquisition of words in the language of lower exposure should, on average, be more susceptible to the effect of cognateness.

In line with this second hypothesis, the vocabulary size associated with linguistic similarity that Floccia et al. reported was larger in the additional language vocabulary than in English vocabulary. Most participants in their sample were English-dominant, meaning that their relative amount of exposure to English was larger than in the additional language. Therefore, participants may have, on average, learnt the English word-form of translation equivalents earlier than the word-form in the additional language.  If this is the case, then the acquisition of English words by English-dominant participants would rarely benefit from their cognate status (the other word-form is not available yet), while the acquisition of words in the additional language would benefit from their phonological similarity with the (available) English form.

In this study, we investigated the role of cognateness on the acquisition of translation equivalent, and whether such effect is conditional to the amount of time the child is exposed to the corresponding language of each member of the translation pair. Using an online vocabulary checklist--designed specifically for this study--we collected data from a sample of children aged `r min(floor(participants$age))` to `r max(floor(participants$age))` months learning Catalan and/or Spanish, with varying degrees of exposure to each language. We then adopted an Item Response Theory (*IRT*) approach to model the probability of participants being reported by their parents to understand or understand *and* say each word in the checklist, conditional to its cognate status in Catalan and Spanish and participants' degree of exposure to the corresponding language, while adjusting for other indices of word difficulty--lexical frequency and length (phonemes)-- and participant skill--age [see @kachergis2021towards for a similar approach].

@@

We collected data from `r printnum(n_participants)` children from the Metropolitan Area of Barcelona between `r format(dates[1], "%dth %B, %Y")` and `r format(dates[2], "%dth %B, %Y")`. All families gave informed consent before participating and this study was approved by the Comitè d'Ètica de la Investigació amb Medicaments (CEIm) from Hospital del Mar (Barcelona, Spain), code XXXXXXXXX. Families from the participant database of the Laboratori de Recerca en Infància of the Universitat Pompeu Fabra were contacted by e-mail or phone if their child were aged between `r floor(min(participants$age))` and `r floor(max(participants$age))` months, and had not been reported to be exposed more than 10% of the time to a language other than Spanish or Catalan. Upon consent, families were sent a link to the questionnaire via e-mail, which they filled from a computer, laptop, or mobile device in a browser within the two week following the invitation to participate. Table 1 summarises the distribution of participants across ages and degrees of exposure (DoE) to Catalan.

```{r tab-participants-lp, message=FALSE, warning=FALSE, paged.print=FALSE, fig.width=7, fig.cap="Participant sample size by age and degree of exposure to Catalan. Information about DoE was provided by families before filling the questionnaire. A 100% indicated that the participant was exclusively exposed to Catalan, and never to Spanish. A 0% indicates that the participant was not exposed to Catalan ever, and rather most of the time to Spanish. A 50% indicates that the participant was exposed to Catalan and Spanish approximately half of the time each. The 100%, 0%, and 50% would be traditionaly classified as Catlaan monolingual, Spanish monolingual, and Catalan-Spanish bilingual, respectively. For illustration purposes this table, DoEs were binned into 25\\%-wide bins, and ages (in months) binned into 4 month-wide bins. Hyphens indicate that no participants from that specific combination of age and DoE filled the questionnaire."}
participants %>% 
    mutate_at(
        vars(starts_with("doe_")), 
        ~floor(.*10)/10
    ) %>% 
    mutate(
        age = cut(
            floor(age),
            c(10, 14, 18, 22, 26, 30, 34, 36),
            include.lowest = TRUE
        ),
        doe_catalan = cut(
            doe_catalan, 
            c(0, 0.25, 0.5, 0.75, 1),
            labels = c("0-25%", "25-50%", "50-75%", "75-100%"),
            include.lowest = TRUE
        )
    ) %>% 
    count(age, doe_catalan) %>% 
    arrange(desc(doe_catalan)) %>% 
    pivot_wider(names_from = age, values_from = n) %>% 
    gt() %>% 
    summary_rows(
        columns = 2:8, 
        fns = list("Total" = ~sum(., na.rm = TRUE)),
        formatter = fmt_integer
    ) %>% 
    sub_missing(everything(), everything(), missing_text = "--") %>% 
    cols_label(doe_catalan = "DoE Catalan") %>%
    tab_spanner("Age (months)", 2:8) %>% 
    tab_footnote(
        "This proportion is complementary to the degree of exposure to Spanish, with the exception of those participants who were also exposed to a third language up to 10% of the time", 
        cells_column_labels(columns = doe_catalan)
    ) %>% 
    tab_style(
        cell_text(align = "left"),
        list(
            cells_body(doe_catalan),
            cells_column_labels(doe_catalan)
        )
    )
```

@@

The questionnaire was implemented on-line using the formR platform [@arslan2020formr], and was structured in three blocks: a (1) language questionnaire, a (2) demographic survey, and a (3) Catalan and a Spanish vocabulary checklists. Vocabulary checklists followed a similar structure as the Oxford Communicative Developmental Inventory [@hamilton2000infant] and consisted in two lists of words: one in Catalan and one in Spanish. The Catalan inventory contained `r n_item_language["catalan"]` items and the Spanish inventory contained `r n_item_language["spanish"]`. Items in one language were translation equivalents of the items in the other language (e.g., whenever *gos* [dog] was included in the Catalan inventory, the word *perro* was included in the Spanish inventory), roughly following a one-to-one mapping. In case two translation equivalents were possible for a given word, both were included as separate items (e.g., Catalan *acabar* [*to finish*] and Spanish *acabar* and *terminar*), or merged them into a single item (e.g., Spanish *mono* [*monkey*] and Catalan *mono/mico*). We included items from a diverse sample of `r n_categories` semantic/functional categories (see Appendix 1). For the analyses included in this study, we excluded items from the adverbs, auxiliary words, connectives, interjections and games and routines categories, so that only data from content words (nouns, adjectives, and verbs) were used.

@@

```{r tab-item-summary, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE, fig.width=7, fig.cap="Lexical properties of the word.forms and trasnlation pairs included in the vocabulary checklist and our analyses. We report the number of items, mean (*M*), standard deviation (*SD*), and range of the lexical frequency (epxressed in Zipf scores), length (number of phonemes) of the word forms, and the phonological similarity (normalised Levenshtein similarity) with their translation equivalent. Statistics are reported separately for Catalan and Spanish, and for the different vocabulary checklists that participants were allocated to (A, B, C, D)."}
items <- left_join(items, select(multilex_data$pool, te, item, language, class, category))

items %>%
    unnest(list) %>% 
    group_by(language, list) %>% 
    summarise_at(
        vars(freq, n_phon, lv), 
        lst(n = ~sum(!is.na(.)), mean, sd, min, max)
    ) %>% 
    select(-c(n_phon_n, lv_n)) %>% 
    rename(n = freq_n) %>% 
    gt(groupname_col = "language") %>% 
    fmt_number(matches("freq_|n_phon_mean|n_phon_sd")) %>% 
    fmt_percent(matches("lv_")) %>% 
    tab_spanner("Frequency (Zipf)", columns = starts_with("freq_")) %>% 
    tab_spanner("# Phonemes", columns = starts_with("n_phon_")) %>% 
    tab_spanner("Phon. similarity (Levenshtein)", columns = starts_with("lv_")) %>% 
    cols_merge_range(col_begin = "freq_min", col_end = "freq_max") %>% 
    cols_merge_range(col_begin = "n_phon_min", col_end = "n_phon_max") %>% 
    cols_merge_range(col_begin = "lv_min", col_end = "lv_max") %>% 
    cols_label(
        list = "",
        n = md("*N*"),
        freq_mean = md("Mean"),
        freq_sd = md("*SD*"),
        freq_min = md("Range"),
        n_phon_mean = md("Mean"),
        n_phon_sd = md("*SD*"),
        n_phon_min = md("Range"),
        lv_mean = md("Mean"),
        lv_sd = md("*SD*"),
        lv_min = md("Range")
    ) %>% 
    tab_style(
        cell_text(style = "italic"),
        cells_column_labels(everything())
    ) %>% 
    tab_style(
        cell_text(weight = "bold"),
        cells_column_spanners(everything())
    )
```

@@
